A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-12 tex/BiSig.tex
[ tex ] Clean up code in S6
[ tex ] Add a separate section on Examples
[ tex ] Minor changes
[ tex ] Revisions up to S6

\documentclass[acmsmall,screen,review,anonymous,draft]{acmart}

@@

% Used for typesetting Agda code
\usepackage{agda-lhs2tex}

@@

\begin{document}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but only on a case-by-case basis.
This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a first theory of bidirectional type synthesis.
The whole theory works as a verified type-synthesiser generator for syntax-directed simple type systems:
From a signature that specifies a bidirectionally typed language, the theory produces a type synthesiser that decides whether an input abstract syntax tree has a typing derivation or reports that the input does not have enough type annotations.
Within the theory, we formally define soundness, completeness, and mode-correctness, which are sufficient conditions for deriving a correct type synthesiser.
We also propose a preprocessing step called `mode decoration', which helps the user to deal with missing type annotations and streamlines the theory.
The entire theory is formulated constructively, and has been formalised in the \Agda proof assistant.

%\LT{Still not coherent enough}
%Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
%In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
%which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
%First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
%which is analogous to a grammar in parsing.
%Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
%Moreover, we introduce `mode decoration' to deal with missing type annotations, an aspect often overlooked in the literature.
%The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
%By being parametrised by a signature, our theory works overall as a verified type-synthesiser generator.
\end{abstract}

@@

\input{sec1-intro}
\input{sec2-outline}
\input{sec3-defs.tex}
\input{sec4-pre-synthesis.tex}
\input{sec5-synthesis.tex}
\subfile{sec6-formalisation}
\input{sec7-future.tex}

>>>>

\documentclass[acmsmall,screen,review,anonymous]{acmart}

@@

% Used for typesetting Agda code
\usepackage{agda-lhs2tex}
\newcommand{\cons}[1]{\mathbf{#1}}
\newcommand{\iden}{\mathit}

@@

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but only on a case-by-case basis.
This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a first theory of bidirectional type synthesis.
The whole theory works as a verified type-synthesiser generator for syntax-directed simple type systems:
From a signature that specifies a simply typed language with a bidirectional type system, the theory produces a type synthesiser that decides whether an input abstract syntax tree has a typing derivation or reports that the input does not have enough type annotations.
Within the theory, we formally define soundness, completeness, and mode-correctness, which are sufficient conditions for deriving a correct type synthesiser.
We also propose a preprocessing step called `mode decoration', which helps the user to deal with missing type annotations and streamlines the theory.
The entire theory is formulated constructively, and has been formalised in the proof assistant \Agda with \AxiomK.

\end{abstract}
\begin{document}

@@

\input{sec1-intro}
\input{sec2-outline}
\input{sec3-defs.tex}
\input{sec4-pre-synthesis.tex}
\input{sec5-synthesis.tex}
\input{example.tex}
\subfile{sec6-formalisation}
\input{sec7-future.tex}

### 2023-07-12 tex/example.tex
[ tex ] Revisions up to S6

%In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
In this language, we introduce additional type constructs apart from $\bto$.
To do so, we extend $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with the operations $\mathsf{nat}$ and $\mathsf{times}$ such that
\[
  \arity({\mathsf{nat}}) = 0 \quad\text{and}\quad \arity({\mathsf{times}}) = 2
\]
enabling the introduction of $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with an operation for each rule above:
\begin{flalign*}
  &  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
  &  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
  && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
  && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
  &  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
  && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
\end{flalign*}
With these operations, a straightforward check shows that every rule is mode-correct, therefore this is a mode-correct bidirectional type system.
As a result, we can instantiate the type synthesiser presented in \Cref{sec:type-synthesis} for this system, eliminating the need for re-implementation.

>>>>

%In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
In this language, we introduce additional type constructs apart from $\bto$.
To do so, we extend $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with the operations $\mathsf{nat}$ and $\mathsf{times}$ such that
\[
  \arity({\mathsf{nat}}) = 0 \quad\text{and}\quad \arity({\mathsf{times}}) = 2
\]
enabling the introduction of $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with an operation for each rule above:
\begin{flalign*}
  &  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
  &  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
  && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
  && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
  &  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
  && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
\end{flalign*}
With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
As a result, we can instantiate the type synthesiser presented in \Cref{sec:type-synthesis} for this system, eliminating the need for re-implementation.

### 2023-07-12 tex/sec1-intro.tex
[ tex ] Revisions up to S6

Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.%
\footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
While there are some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type theories.
Moreover, their algorithms are not proved complete.}
%While it is straightforward to derive a type synthesis algorithm,
For bidirectional typing, \citet{Dunfield2021} could only present informal design principles that the community learned from individual systems, rather than a general theory providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
To remedy the situation, we initiate a theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed simple type systems.

@@

This paper is structured as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The most heavyweight technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
We briefly sketch the design of our formalisation in \cref{sec:formalisation} and conclude in \cref{sec:future} with potential challenges in extending our theory to a more general setting.

>>>>

Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools, notably parser generators.%
\footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
While there are some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type theories.
Moreover, their algorithms are not proved complete.}
%While it is straightforward to derive a type synthesis algorithm,
For bidirectional typing, \citet{Dunfield2021} could only present informal design principles that the community learned from individual systems, rather than a general theory providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
To remedy the situation, we initiate a theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed simple type systems.

@@

This paper is structured as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The most heavyweight technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
We briefly sketch our \Agda formalisation in \cref{sec:formalisation}.
Finally, in \cref{sec:future} we reflect on techniques that may improve the formal development and potential challenges in extending our theory to a more general setting.

### 2023-07-12 tex/sec2-outline.tex
[ tex ] Revisions up to S6

\begin{definition}
\label{def:STLC-type-synthesiser}
Parametrised by an `excuse' predicate~$E$ on raw terms, a \emph{type synthesiser} takes a context~$\Gamma$ and a raw term $\erase\Gamma \vdash t$ (where $\erase\Gamma$~is the domain of~$\Gamma$) as input, and establishes one of the following outcomes:
\begin{enumerate}
\item there exists a derivation of $\Gamma \vdash t : A$ for some type~$A$,
\item there does not exist a derivation $\Gamma \vdash t : A$ for any type~$A$, or
\item $E$~holds for~$t$.
\end{enumerate}
\end{definition}

@@

While it is possible for a bidirectional type synthesiser to do its job in one go, which can be thought of as adding both mode and typing information to a raw term and arriving at a bidirectional typing derivation, it is beneficial to have a preprocessing step which adds only mode information, based on which the synthesiser then continues to add typing information.
More precisely, the preprocessing step attempts to produce \emph{mode derivations} as defined in \cref{fig:STLC-mode-derivations}, where the rules are exactly the mode part of the bidirectional typing rules~(\cref{fig:STLC-bidirectional-typing-derivations}).

>>>>

\begin{definition}
\label{def:STLC-type-synthesiser}
Parametrised by an `excuse' predicate~$E$ on raw terms, a \emph{type synthesiser} takes a context~$\Gamma$ and a raw term $\erase\Gamma \vdash t$ (where $\erase\Gamma$~is the list of variables in~$\Gamma$) as input, and establishes one of the following outcomes:
\begin{enumerate}
\item there exists a derivation of $\Gamma \vdash t : A$ for some type~$A$,
\item there does not exist a derivation $\Gamma \vdash t : A$ for any type~$A$, or
\item $E$~holds for~$t$.
\end{enumerate}
\end{definition}

@@

While it is possible for a bidirectional type synthesiser to do its job in one go, which can be thought of as adding both mode and typing information to a raw term and arriving at a bidirectional typing derivation, it is beneficial to have a preprocessing step which adds only mode information, based on which the synthesiser then continues to add typing information.
More precisely, the preprocessing step, which we call \emph{mode decoration}, attempts to produce \emph{mode derivations} as defined in \cref{fig:STLC-mode-derivations}, where the rules are exactly the mode part of the bidirectional typing rules~(\cref{fig:STLC-bidirectional-typing-derivations}).

### 2023-07-12 tex/sec3-defs.tex
[ tex ] Revisions up to S6

\begin{definition}\label{def:typing-derivations}
  \begin{figure}
    \centering
    \small
    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a concrete type $A$ under $\Gamma$ for a simply typed language $(\Sigma, \Omega)$}
    \begin{mathpar}
      \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
      \and
      \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
      {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
      \\
    \text{for $\bop$ in $\Omega$}
    \end{mathpar}
    \caption{Typing derivations}
    \label{fig:extrinsic-typing}
  \end{figure}
  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$, indexed by a typing context $\Gamma : \Cxt_\Sigma$, a raw term~$\isTerm{t}$ with free variables in $\erase{\Gamma}$, and a type $A : \Type_\Sigma(\emptyset)$, consists of 
  \begin{enumerate}
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} x : A$ if $x : A$ is in $\Gamma$,
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\bop$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
  \end{enumerate}
  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
\end{definition}

@@

\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
  \begin{figure}
    \centering
    \small
    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{$\isTerm{t}$ has a concrete type $A$ in mode $\dir{d}$ under $\Gamma$ for a bidirectional type system $(\Sigma, \Omega)$} 
    \begin{mathpar}
      \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ B = A}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
      \\
      \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
        \cdots \\
      \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
      {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
      \\ \text{for $\biop$ in $\Omega$}
    \end{mathpar}
    \caption{Bidirectional typing derivations}
    \label{fig:bidirectional-typing-derivations}
  \end{figure}
  \begin{figure}
    \centering
    \small
    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is in mode $d$ with free variables in $V$ for a bidirectional type system $(\Sigma, \Omega)$}
    \begin{mathpar}
      \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
      \and
      \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
      \and
      \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
      \and
      \inferrule{V, \vars{x}_1 |-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \\ \cdots \\ V, \vars{x}_{n} |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
      {V |-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
      \and \text{for $\biop$}
    \end{mathpar}
    \caption{Mode derivations}
    \label{fig:mode-derivations}
  \end{figure}
  For a bidirectional type system $(\Sigma, \Omega)$,
  \begin{itemize}
    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
          \[
            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
          \]
          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
  \end{itemize}
  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
\end{definition}

>>>>

\begin{definition}\label{def:typing-derivations}
  \begin{figure}
    \centering
    \small
    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a concrete type $A$ under $\Gamma$ for a simply typed language $(\Sigma, \Omega)$}
    \begin{mathpar}
      \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
      \and
      \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
      {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
      \\
    \text{for $\bop$ in $\Omega$}
    \end{mathpar}
    \caption{Typing derivations}
    \label{fig:extrinsic-typing}
  \end{figure}
  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$, indexed by a typing context $\Gamma : \Cxt_\Sigma$, a raw term~$\isTerm{t}$ with free variables in $\erase{\Gamma}$, and a type $A : \Type_\Sigma(\emptyset)$, consists of 
  \begin{enumerate}
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} x : A$ if $x : A$ is in $\Gamma$,
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\bop$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
  \end{enumerate}
  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
\end{definition}

@@

\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
  \begin{figure}
    \centering
    \small
    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{$\isTerm{t}$ has a concrete type $A$ in mode $\dir{d}$ under $\Gamma$ for a bidirectional type system $(\Sigma, \Omega)$} 
    \begin{mathpar}
      \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ B = A}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
      \\
      \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
        \cdots \\
      \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
      {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
      \\ \text{for $\biop$ in $\Omega$}
    \end{mathpar}
    \caption{Bidirectional typing derivations}
    \label{fig:bidirectional-typing-derivations}
  \end{figure}
  \begin{figure}
    \centering
    \small
    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is in mode $d$ with free variables in $V$ for a bidirectional type system $(\Sigma, \Omega)$}
    \begin{mathpar}
      \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
      \and
      \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
      \and
      \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
      \and
      \inferrule{V, \vars{x}_1 |-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \\ \cdots \\ V, \vars{x}_{n} |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
      {V |-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
      \and \text{for $\biop$}
    \end{mathpar}
    \caption{Mode derivations}
    \label{fig:mode-derivations}
  \end{figure}
  For a bidirectional type system $(\Sigma, \Omega)$,
  \begin{itemize}
    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
          \[
            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
          \]
          has a derivation for $\biop$ in $\Omega$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
  \end{itemize}
  {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} are shorthands for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ respectively.
  A typing rule is \emph{checking} if its conclusion mode is $\chk$ and \emph{synthesising} if its conclusion mode is $\syn$.
\end{definition}

### 2023-07-12 tex/sec4-pre-synthesis.tex
[ tex ] Revisions up to S6

The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better:
If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
\footnote{For brevity, we use ordinary mode decoration elsewhere in this paper.}

@@

The theorem could be proved directly, but that would mix up two case analyses which respectively inspect the input term and apply mode casts depending on what modes are required.
Instead, we distill the case analysis that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:strict-derivation}) is then established by induction on the input term in the proof of \cref{thm:generalised-mode-decoration}.

@@

\citet{Dunfield2021} formulated completeness differently from ours (\cref{thm:STLC-completeness,sec:soundness-and-completeness}) and proposed \emph{annotatability} as a more suitable name.
In our theory, we may formulate annotatability as follows.

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be easily proved from generalised mode decoration.

>>>>

The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better:
If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
\footnote{To simplify the presentation, we use ordinary mode decoration elsewhere in this paper.}

@@

The theorem could be proved directly, but that would mix up two case analyses which respectively inspect the input term~$\isTerm{t}$ and apply mode casts depending on which mode~$\dir{d}$ is required.
Instead, we distill the case analysis on~$\dir{d}$ that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:strict-derivation}) is then established by induction on~$\isTerm{t}$ in the proof of \cref{thm:generalised-mode-decoration}.

@@

\citet[Section~3.2]{Dunfield2021} formulated completeness differently from ours (\cref{thm:STLC-completeness,sec:soundness-and-completeness}) and proposed \emph{annotatability} as a more suitable name.
In our theory, we may formulate annotatability as follows.

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be easily proved with generalised mode decoration.
